Modelling BCG vaccination in the UK

### 2019-05-03 index/01-introduction.Rmd
reviewed introduction

The BCG vaccine was developed in 1921 and was introduced to the UK in 1953. Globally, it has been shown to offer variable protection that may reduce over time.[@TheWorldHealthOrganization:2018va] However, there is strong evidence that BCG offers high levels of protection for children, and more generally within the UK born population.[@Roy2014] It remains the only licensed TB vaccine with over 100 million doses given globally each year. Serious side effects are very rare but scarring commonly occurs at the site of injection. In 2005, the UK withdrew the universal BCG program for those at school age and introduced a targeted program of vaccination for babies that were deemed to be at high risk.[@Zwerling2011a] This was motivated by several years of declining transmission, the evidence of high levels of protection in children and a belief that other control measures would be more cost-effective.[@Zwerling2011a] Since this change in policy, declining incidence appears to support this decision.[@PHE2017] However, due to TB’s complex dynamics, the long-term effects are difficult to predict.

@@

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. I makes use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. I then use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objects that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

In this thesis, data visualisation is used extensively in Chapter \@ref(background) to explore the epidemiology of TB globally and in Chapter \@ref(data) to explore the epidemiology of TB in England. Chapter \@ref(data) also uses visualisation to generate many of the hypotheses that are then explored in further detail throughout the rest of this thesis. The remaining thesis chapters use data visualisation to explore the data used in the analysis and the results from the analysis.

@@

In this thesis, a partially observed stochastic compartmental model of Tuberculosis is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

  - To describe the current epidemiology of Tuberculosis in England, in the context of global Tuberculosis epidemiology.
  - To assess some of the statistical modelling evidence used to justify the 2005 change in BCG vaccination policy in the UK.
  - To assess whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for TB cases in England.
  - To assess the effects of the 2005 change in vaccination policy on those eligible for vaccination.
  - To develop a parsimonious transmission dynamic model of tuberculosis which captures current, and historic. vaccination policy and reflects our current understanding of TB epidemiology in England.
  - To fit this model using all available data sources.
  - To investigate the effectiveness of universal school-age vs. universal neonatal vs. no vaccination using the previously developed transmission dynamic model.

@@

* **Chapter \@ref(sutherland):** This chapter recreates a simulation based statistical model that was used as part of the decision making process that led to the 2005 change in BCG vaccination policy. It extends the previously implemented model by fully capturing all parameter and model uncertainty, and updating the underlying data. It then estimates the impact in real-terms of the change in policy using this updated model.

@@

* **Chapter \@ref(model-development):** In this chapter a mechanistic model of TB and BCG vaccination in England is developed. The model structure is fully justified based on the known epidemiology of TB in England. Model parameters are given prior distributions based on routine surveillance data (Chapter \@ref(data)), the published literature, and assumptions based on expert knowledge where no other source exists.

* **Chapter \@ref(model-fitting):** In this chapter the model developed in the previous chapter is fitted to the routine surveillance data (Chapter \@ref(data)) using bayesian methods. Multiple scenarios are considered, with the results evaluated using their DIC.

@@

This thesis has produced both traditional and non-traditional output. The traditional output includes peer reviewed papers, preprints and talks at academic conferences. The non-traditional output includes open source research software, software for improving the academic workflow, dashboards for exposing relevant data, dashboards for exploring the modelling methods used in this thesis, and an educational dashboard for teaching some the benefits of vaccination. These outputs are detailed in the following section.

@@

* **`getTBinR`**: The `getTBinR` R package facilitates downloading the most up-to-date version of multiple Tuberculosis relevant data sources from the World Health Organization, along with the accompanying data dictionaries. It also contains functions to allow easy exploration of the data via searching data dictionaries, summarising key metrics on a regional and global level, and visualising the data in a variety of highly customisable ways. See Chapter \@ref(getTBinR) for further details. Install from CRAN with `install.packages("getTBinR")` or install the development version from GitHub with `devtools::install_github("seabbs/getTBinR")`. Link: https://www.samabbott.co.uk/getTBinR/

* **`tbinenglanddataclean`**: An R package that contains the functions and documentation required to reproduce all data import and munging used in this thesis. This package provides a workflow to facilitate reproducing all analysis in this thesis and  to expedite the work of others using data from the Enhanced Surveillance System (Chapter \@ref(data)). Available from GitHub using `devtools::install_github("seabbs/tbinenglanddataclean")`. Link: https://www.samabbott.co.uk/tbinenglanddataclean/

* **`idmodelr`**: An R package that contains a library infectious disease models as well as modelling utilities. It provides tooling that includes: example SEI/SEIR/SHLIR/SHLITR model code, a model solving wrapper; a model summary function; a scenario analysis function). Used by the Explore infectious disease model dashboard (http://seabbs.co.uk/shiny/exploreidmodels/) for all functionality. Available from GitHub using `devtools::install_github("seabbs/idmodelr")`. Link: https://www.samabbott.co.uk/idmodelr/

@@

- This chapter provides an introduction to Tuberculosis and the BCG vaccine. It then motivates the remainder of this thesis. 
- An outline of the theoretical framework used throughout this thesis is given.
- The aims and objectives of this thesis are detailed.
- An overview of the chapters is provided.
- Finally the dissemination of this work so far is summarised, broken down into peer reviewed output, preprints, software output and talks given at academic conferences.

>>>>

The BCG vaccine was developed in 1921 and was introduced to the UK in 1953. Globally, it has been shown to offer variable protection that may reduce over time.[@TheWorldHealthOrganization:2018va] However, there is strong evidence that BCG offers high levels of protection for children, and more generally within the UK born population.[@Roy2014] It remains the only licensed TB vaccine with over 100 million doses given globally each year. Serious side effects are rare but scarring commonly occurs at the site of injection. In 2005, the UK withdrew the universal BCG program for those at school age and introduced a targeted program of vaccination for babies that were deemed to be at high risk.[@Zwerling2011a] This was motivated by several years of declining transmission, the evidence of high levels of protection in children and a belief that other control measures would be more cost-effective.[@Zwerling2011a] Since this change in policy, declining incidence appears to support this decision.[@PHE2017] However, due to TB’s complex dynamics, the long-term effects are difficult to predict.

@@

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. As a first step I outline some of the key background information and discuss the tooling that I developed in order to explore this information more fully. I then make use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. Next, I use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objects that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

In this thesis, data visualisation is used in Chapter \@ref(background) to explore the epidemiology of TB globally and in Chapter \@ref(data) to explore the epidemiology of TB in England. Chapter \@ref(data) also uses visualisation to generate many of the hypotheses that are then explored in further detail throughout the rest of this thesis. The remaining thesis chapters use data visualisation to explore data and results.

@@

In this thesis, a partially observed stochastic compartmental model of Tuberculosis is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

  - To describe the current epidemiology of Tuberculosis in England, in the context of global Tuberculosis epidemiology.
  - To assess some of the statistical modelling evidence used to justify the 2005 change in BCG vaccination policy in the UK.
  - To assess whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for TB cases in England.
  - To assess the effects of the 2005 change in vaccination policy on those eligible for vaccination.
  - To develop a parsimonious transmission dynamic model of tuberculosis which captures current, and historic, vaccination policy and reflects our current understanding of TB epidemiology in England.
  - To fit this model using all available data sources.
  - To investigate the effectiveness of universal school-age vs. universal neonatal vs. no vaccination using the previously developed transmission dynamic model.

@@

* **Chapter \@ref(sutherland):** This chapter recreates a simulation based statistical model that was used as part of the decision making process that led to the 2005 change in BCG vaccination policy. It extends the previously implemented model by capturing parameter and model uncertainty, and updating the underlying data. It then estimates the impact in real-terms of the change in policy using this updated model.

@@

* **Chapter \@ref(model-development):** In this chapter a mechanistic model of TB and BCG vaccination in England is developed. The model structure is justified based on the known epidemiology of TB in England. Model parameters are given prior distributions based on routine surveillance data (Chapter \@ref(data)), the published literature, and assumptions based on expert knowledge where no other source exists.

* **Chapter \@ref(model-fitting):** In this chapter the model developed in the previous chapter is fitted to the routine surveillance data (Chapter \@ref(data)) using bayesian methods. Multiple scenarios are considered.

@@

This thesis has produced: peer reviewed papers; preprints; talks at academic conferences; open source research software; open source software for improving the academic workflow; dashboards for exposing relevant data; dashboards for exploring the modelling methods used in this thesis; and an educational dashboard for teaching some the benefits of vaccination. These outputs are detailed in the following section.

@@

* **`getTBinR`**: The `getTBinR` R package facilitates downloading the most up-to-date version of multiple Tuberculosis relevant data sources from the World Health Organization, along with the accompanying data dictionaries. It also contains functions to allow easy exploration of the data via searching data dictionaries, summarising key metrics on a regional and global level, and visualising the data in a variety of customisable ways. See Chapter \@ref(getTBinR) for further details. Install from CRAN with `install.packages("getTBinR")` or install the development version from GitHub with `devtools::install_github("seabbs/getTBinR")`. Link: https://www.samabbott.co.uk/getTBinR/

* **`tbinenglanddataclean`**: An R package that contains the functions and documentation required to reproduce all data import and munging used in this thesis. This package provides a workflow to facilitate reproducing all analyses in this thesis and to expedite the work of others using data from the Enhanced Surveillance System (Chapter \@ref(data)). Available from GitHub using `devtools::install_github("seabbs/tbinenglanddataclean")`. Link: https://www.samabbott.co.uk/tbinenglanddataclean/

* **`idmodelr`**: An R package that contains a library  of infectious disease models as well as modelling utilities. It provides tooling that includes: example SEI/SEIR/SHLIR/SHLITR model code, a model solving wrapper; a model summary function; a scenario analysis function). Used by the explore infectious disease model dashboard (http://seabbs.co.uk/shiny/exploreidmodels/) for all functionality. Available from GitHub using `devtools::install_github("seabbs/idmodelr")`. Link: https://www.samabbott.co.uk/idmodelr/

@@

- This chapter provides an introduction to Tuberculosis and the BCG vaccine. It then motivates the remainder of this thesis. 
- An outline of the theoretical framework used throughout this thesis is given.
- The aims and objectives of this thesis are detailed.
- An overview of the chapters is provided.
- Finally the dissemination of this work so far is summarised, broken down into peer reviewed output, preprints, software output, and talks given at academic conferences.
